\documentclass[12pt,a4paper]{article}

% ─── Packages ─────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{array}
\usepackage{longtable}
\usepackage{parskip}
\usepackage{fancyhdr}
\usepackage{lastpage}

% ─── Page geometry ────────────────────────────────────────────
\geometry{margin=2.5cm}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small CSC2106 --- Real-Time File Transfer System using CoAP}
\fancyhead[R]{\small BugattiFerrariLamborghini}
\fancyfoot[C]{\thepage\ of \pageref{LastPage}}

% ─── Section formatting ──────────────────────────────────────
\titleformat{\section}{\large\bfseries}{Section \Roman{section}: }{0em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{1em}{}

% ─── Hyperref ─────────────────────────────────────────────────
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}

% ─── Custom commands ──────────────────────────────────────────
\newcommand{\changed}[1]{\textcolor{red}{#1}}
\newcommand{\forecast}[1]{\textcolor{red}{\textbf{[Forecasted]} #1}}

\begin{document}

% ═════════════════════════════════════════════════════════════
% TITLE PAGE
% ═════════════════════════════════════════════════════════════

\begin{center}
    \vspace*{1cm}
    {\Large \textbf{Project: [CSC2106] Real-Time File Transfer System using CoAP}} \\[0.5cm]
    {\large Date: 18 March 2026} \\[0.3cm]
    {\large Team Name: BugattiFerrariLamborghini} \\[1cm]

    \begin{tabular}{|p{7cm}|p{4cm}|}
        \hline
        \textbf{Student Name} & \textbf{Student ID} \\
        \hline
        Pang Jia De & 2402226 \\
        \hline
        Dui Rui En Joshua & 2402201 \\
        \hline
        Tan Jian Xin & 2401421 \\
        \hline
        Tan Bing Kun Terence & 2401883 \\
        \hline
    \end{tabular}
\end{center}

\vspace{1.5cm}

% ═════════════════════════════════════════════════════════════
% ABSTRACT
% ═════════════════════════════════════════════════════════════

\section*{Abstract}
% (150--250 words)

Environmental monitoring in remote, off-grid locations faces a critical ``data latency gap'' due to the physical limitations of wireless sensor networks. Traditional systems force a compromise between high bandwidth, which rapidly depletes batteries, and long-range, low-power communication, which cannot support heavy payloads like image data. This project resolves this ``Bandwidth-Range-Power Trilemma'' by developing a Hybrid IoT Mesh Network built on ESP32-S3 microcontrollers \changed{with LILYGO T3-S3 V1.2 boards featuring the SX1280 2.4\,GHz LoRa transceiver}. The proposed architecture decouples the control and data planes: a LoRa radio module serves exclusively as a low-power control plane, utilizing hardware interrupts to keep sensor nodes in deep sleep until an event triggers a wake-up. Once activated, the nodes \changed{each run an independent Wi-Fi access point; the gateway connects as a station (STA) to harvest images via CoAP block-wise transfer over UDP}. High-bandwidth payloads, simulated in this prototype via pre-loaded images on SD cards, are segmented and \changed{reliably} transmitted \changed{over the Wi-Fi data plane} using CoAP block-wise transfer over UDP. \changed{The system employs FreeRTOS dual-core task separation to eliminate LoRa blackouts during harvest, AODV reactive mesh routing for multi-hop coordination, a Bully election algorithm for automatic gateway selection, and automatic role negotiation so nodes self-organize without manual configuration.} By shifting from a continuous-listening network to an event-driven hybrid model, this system drastically minimizes idle power consumption while retaining the capacity to route large files across multiple hops. The resulting prototype demonstrates a scalable, energy-efficient framework capable of delivering real-time visual alerts to a local gateway, shifting remote forest monitoring from reactive manual data retrieval to proactive, real-time intervention.

\newpage

% ═════════════════════════════════════════════════════════════
% SECTION I: INTRODUCTION, PROBLEM STATEMENT AND OBJECTIVES
% ═════════════════════════════════════════════════════════════

\section{Introduction, Problem Statement and Objectives}
% (Word Limit: 750 words)

\subsection{Introduction}

Effective environmental monitoring and forest conservation is critical for biodiversity preservation and disaster mitigation, which require sensor networks capable of operating autonomously in remote, off-grid environments. Currently, the industry relies heavily on manual data retrieval from local storage (Meek et al., 2015) [1], which is a labor-intensive and reactive approach. This project proposes a Hybrid IoT Mesh Network utilizing ESP32-S3 microcontrollers to bridge the gap between remote sensing and real-time data access. By integrating LoRa for long-range, low-power control signaling and Wi-Fi for high-bandwidth data transfer, the system offers a highly scalable solution. \changed{To ensure rigorous and deterministic validation of the underlying network routing, this iteration of the project replaces live camera hardware with an SD card containing pre-loaded images. This approach utilizes images with various resolutions pre-loaded onto SD cards, eliminating camera hardware-induced errors and isolating the network's performance metrics for more accurate testing.} Once the network is awakened, all high-bandwidth communications utilize CoAP over UDP to enable lightweight, reliable image transmission via block-wise transfer, shifting remote monitoring from reactive data retrieval to proactive intervention.

\subsection{Problem Statement}

The primary challenge in off-grid environmental monitoring is the ``Data Latency Gap.'' Standard methods store data locally on SD cards, resulting in weeks or months of delay between an event occurring and its detection. While real-time wireless alternatives exist, they face a strict ``Bandwidth-Range-Power Trilemma'': low-power wide-area technologies (like LoRa) lack the bandwidth required for image transfer, while high-bandwidth solutions (like standard Wi-Fi or cellular networks) consume excessive energy and lack coverage in deep forests.

This project directly addresses this gap by decoupling the control and data transport layers, solving the power constraints that traditionally limit the long-term deployment of visual sensors. \changed{Furthermore, traditional embedded vision nodes frequently suffer from heap memory exhaustion when processing large image buffers in real-time. By implementing a storage-driven data pipeline that streams data directly from an SD card to the Wi-Fi buffer in 1024-byte chunks, this project circumvents the RAM limitations of the ESP32-S3. Additionally, the network design has been improved to operate completely independently of commercial LoRaWAN or 4G infrastructure; the gateway is now implemented purely as a designated ``Root Node'' within the network, bridging directly to a local diagnostic server for highly localized, off-grid deployment.}

\subsection{Objectives}

The main goal of this project is to develop and validate a functional prototype of a hybrid sensor node that overcomes the limitations of single-protocol networks. The specific objectives are to:

\begin{enumerate}
    \item \textbf{Design a Hybrid Network Architecture:} Develop a sensor node utilizing the ESP32-S3 that integrates both LoRa \changed{(SX1280)} and Wi-Fi radios to optimize the trade-off between transmission range and data bandwidth.

    \item \textbf{Implement Wake-on-Radio Coordination:} Engineer a highly efficient power management system using LoRa-based hardware interrupts (DIO1) to maintain nodes in a deep sleep state ($<$100\,$\mu$A) until triggered by a network event.

    \item \textbf{Optimize Image Transfer:} Implement CoAP over UDP with Block-Wise Transfer mechanisms to enable the reliable transmission of large payloads over a lossy, multi-hop Wi-Fi mesh.

    \item \textbf{Develop Dynamic Role Assignment:} Implement a flexible role configuration system allowing all ESP32-S3 nodes to be assigned as Leaf, Relay, or Gateway at deployment, with persistence across power cycles. \changed{Extend this with automatic role negotiation via a Bully election algorithm, enabling nodes to self-organize without manual configuration.}

    \item \textbf{Implement Queue-Based Scheduling:} Design a gateway-managed transmission queue that assigns time-division slots to prevent network congestion, ensuring only one node transmits image data at a time.

    \item \changed{\textbf{Develop a File Reading Pipeline:} Design an SD-card-based file I/O system that simulates physical camera captures by randomly reading and segmenting pre-loaded JPEG files into network-safe chunks, ensuring stable payload generation without memory allocation failures.}

    \item \textbf{Evaluate System Performance:} Quantitatively assess the prototype against standard engineering metrics, specifically measuring idle power consumption, total network throughput, and successful multi-hop packet delivery rates in a simulated deployment environment.
\end{enumerate}

\newpage

% ═════════════════════════════════════════════════════════════
% SECTION II: LITERATURE REVIEW
% ═════════════════════════════════════════════════════════════

\section{Literature Review}
% (Word Limit: 500 words)

\subsection{Current State of Knowledge and Technology}

The current state of environmental monitoring is transitioning from offline, manual camera traps to connected Wireless Sensor Networks (WSNs) [2], [3]. However, existing deployments remain highly fragmented based on radio capabilities. High-bandwidth networks generally rely on 4G/LTE or standard Wi-Fi, which lack coverage in remote forests and require substantial power infrastructure [4]. Conversely, Low-Power Wide-Area Networks (LPWANs) like LoRa provide excellent multi-kilometer range and run on small batteries, but are strictly limited to low-byte telemetry [5].

\subsection{Significant Research Findings and Technological Developments}

Recent technological developments have heavily focused on optimizing routing protocols and energy management. Studies demonstrate that LoRa excels in ultra-low-power sensing, capable of sustaining nodes for years [6]. \changed{Advancements in AODV (Ad-hoc On-Demand Distance Vector) routing have enabled reactive, infrastructure-less topologies capable of multi-hop coordination without centralized management [7].} The most significant finding in recent hybrid-network literature is the inherent inadequacy of single-radio systems to simultaneously handle long-range, continuous idle listening and high-speed data bursts efficiently [8].

\subsection{Common Methodologies and Their Effectiveness}

Commonly used methodologies in previous studies involve single-radio duty-cycling, where a high-bandwidth node (Wi-Fi/Cellular) wakes up on a fixed schedule [9]. While this extends battery life, it is largely ineffective for real-time alerts because the network is ``deaf'' to incoming commands between cycles. Another attempted methodology is transmitting heavily compressed image data exclusively over LoRa. This has proven highly ineffective; transmitting a standard JPEG over LoRa occupies the frequency for several minutes, causing severe network congestion, high packet loss, and rapid battery depletion during the prolonged transmission window [10].

\subsection{Limitations and Shortcomings of Existing Solutions}

The primary limitation of existing solutions is the ``Bandwidth-Range-Power Trilemma.'' High-bandwidth radios consume excessive idle power, while low-power radios cannot physically handle media files without congesting the network. Furthermore, existing embedded vision methodologies frequently overlook hardware memory constraints. Traditional architectures attempt to load entire image buffers into the microcontroller's RAM before transmission, frequently leading to heap exhaustion, memory fragmentation, and system crashes on constrained devices like the ESP32.

\subsection{Relation of Previous Research to the Objectives of this Study}

The disadvantages identified in previous research lead to the objectives of this study. Recognizing the failure of single-radio architectures, this project adopts a dual-radio, decoupled approach. It leverages the proven capabilities of LoRa strictly as an asynchronous ``Wake-on-Radio'' control plane to trigger out-of-band Wi-Fi mesh data transfers, minimising idle power. Additionally, to overcome the memory exhaustion limitations identified in existing embedded research, this study implements a storage-driven file reading pipeline, acting as images taken by the camera. By utilizing CoAP block-wise transfer to stream image payloads in 1024-byte chunks directly from an SD card through the network, this methodology bypasses RAM limitations, fulfilling the objective of creating a highly reliable, power-efficient, multi-hop image transmission system for off-grid environments.

\newpage

% ═════════════════════════════════════════════════════════════
% SECTION III: DESIGN
% ═════════════════════════════════════════════════════════════

\section{Design}
% (Word Limit: 750 words)

\subsection{Overall System Architecture}

The system employs a decoupled, dual-radio hybrid architecture designed specifically for off-grid forest environments. The network is divided into two distinct operational planes: a low-power ``Control Plane'' for asynchronous wake-ups and a high-bandwidth ``Data Plane'' for payload routing.

\subsection{Hardware and Devices}

The network is composed of identical hardware nodes acting in different roles (Leaf Nodes, Relay Nodes, and one Root Gateway Node). Each node utilizes:

\begin{enumerate}
    \item \textbf{Microcontroller:} Espressif ESP32-S3 (managing logic and Wi-Fi).
    \item \textbf{Transceiver:} \changed{SX1280 2.4\,GHz LoRa module on the LILYGO T3-S3 V1.2 board, connected via FSPI (handling the Control Plane). The board integrates the LoRa PA (Power Amplifier) with RXEN/TXEN switching on dedicated GPIOs.}
    \item \textbf{Display:} SSD1306 128$\times$64 OLED via I2C for real-time status (role, IP, image count, network stats).
    \item \textbf{File Transfer:} A MicroSD Card module via HSPI. To ensure deterministic testing and prevent memory exhaustion, the project simulates camera hardware. High-resolution environmental images (JPEGs) are pre-loaded onto the SD card to act as file transfer data.
\end{enumerate}

\subsection{Network Topology and Connections}

\changed{During the idle phase, all nodes communicate solely via LoRa Peer-to-Peer (P2P). Each leaf and relay node runs its own Wi-Fi access point (AP); the gateway connects as a Wi-Fi station (STA) to each node's AP in sequence during harvest. Multi-hop coordination is handled via AODV reactive routing over the LoRa control plane rather than ESP-Mesh-Lite.}

When a harvest event is triggered, the gateway:
\begin{enumerate}
    \item Sends a \texttt{WAKE\_PING} over LoRa to wake sleeping nodes via DIO1 interrupt.
    \item Waits 2.5\,s for the target node to reinitialize its Wi-Fi AP.
    \item Connects as a station to the node's AP.
    \item Downloads images via CoAP Block-Wise Transfer.
    \item Disconnects and repeats for the next node.
\end{enumerate}

For multi-hop nodes (out of direct Wi-Fi range), the gateway commands a relay over LoRa to perform the harvest via a store-and-forward \texttt{HARVEST\_CMD}/\texttt{HARVEST\_ACK} protocol.

\subsection{Data Collection Pipeline}

Data collection is driven by a custom storage-to-network pipeline. Instead of loading an entire image into the microcontroller's fragile RAM, the ESP32-S3 reads the pre-loaded JPEG from the SD card in chunks of 1024 bytes.

\subsection{Communication Protocols}

\begin{enumerate}
    \item \textbf{Control Protocol:} \changed{LoRa 2.4\,GHz physical layer is used for beacon discovery (31-byte packets every 30\,s), AODV routing (RREQ/RREP/RERR), election packets, wake-up pings, and harvest commands. A 3-byte magic header (\texttt{0xFC 0x01 type}) identifies ForestCam packets.}
    \item \textbf{Transport Protocol:} Data routing over the active Wi-Fi connection relies on CoAP (Constrained Application Protocol) over UDP.
    \item \textbf{Data Transfer Mechanism:} To handle large image files over lossy networks, the system implements CoAP Block-Wise Transfer (RFC 7959). The 1024-byte image chunks from the SD card are mapped perfectly to CoAP block sizes, ensuring they fit safely within the standard 1500-byte Wi-Fi MTU to prevent IP fragmentation. \changed{Pipelined requests with a window of 3 and a 2-second timeout further improve throughput.}
\end{enumerate}

\subsection{Data Storage and Processing}

Data storage is decentralized at the edge. The raw environmental data (simulated images) resides permanently on the local SD cards of the individual leaf nodes. \changed{Images are never deleted from the source node; harvesting is a copy operation. The gateway copies its own \texttt{/images/} to \texttt{/received/} locally and saves harvested images with unique filenames incorporating node MAC suffix, boot count, and uptime (e.g., \texttt{node\_AABB\_boot005\_003672s\_img\_000.jpg}).}

Heavy computational tasks, such as image reconstruction, compression, or full-file buffering, are strictly shifted away from the constrained microcontrollers to prevent heap memory exhaustion. The Root Node (Gateway) receives the CoAP blocks, temporarily buffers them, and streams them via a local Wi-Fi hotspot to a connected diagnostic laptop.

\subsection{User Interaction and Integration}

As this system is designed for highly remote, off-grid environmental monitoring, it strictly avoids integration with third-party cloud platforms. Instead, user interaction occurs through a Local Diagnostic Server. A user (e.g., a forest ranger) enters the physical vicinity of the Root Node with a laptop. \changed{Additionally, serial commands (\texttt{block}, \texttt{unblock}, \texttt{list}) allow runtime management of which nodes participate in harvest cycles.}

\subsection{Security Protocols}

To protect the IoT devices and the integrity of the environmental data, security is implemented at two network layers:

\begin{enumerate}
    \item \textbf{Mesh Layer Security:} The Wi-Fi data plane is secured using standard WPA2-PSK (Wi-Fi Protected Access 2) encryption, preventing unauthorized devices from intercepting traffic.
    \item \textbf{Control Layer Security:} The LoRa wake-up signals are vulnerable to replay attacks. \changed{AES-128 encryption for LoRa packets is planned but not yet implemented. The current prototype relies on a 3-byte magic header for packet identification, which is adequate for controlled laboratory testing.}
\end{enumerate}

\newpage

% ═════════════════════════════════════════════════════════════
% SECTION IV: IMPLEMENTATION (PROTOTYPE AND TESTING)
% ═════════════════════════════════════════════════════════════

\section{Implementation (Prototype and Testing)}
% (Word Limit: 750 words)

\subsection{Prototyping and Development Tools}

The system is prototyped using a modular, staggered approach to isolate hardware and software variables.

\begin{enumerate}
    \item \textbf{Hardware Platforms:} \changed{The core processing units are LILYGO T3-S3 V1.2 development boards integrating the ESP32-S3 and SX1280 2.4\,GHz LoRa with PA. Each board has an SSD1306 OLED display and a MicroSD card slot connected via HSPI (MOSI=11, MISO=2, CLK=14, CS=13). The LoRa transceiver uses FSPI (MOSI=6, MISO=3, SCK=5, CS=7, DIO1=9, BUSY=36, RST=8, RXEN=21, TXEN=10).}

    \item \textbf{Software Development Environment:} Development is conducted in C++ using the PlatformIO IDE (integrated into Visual Studio Code). PlatformIO provides advanced dependency management and multi-file project structuring, strictly required for a multi-developer team working simultaneously on the LoRa, Mesh, and Storage modules.

    \item \textbf{Core Libraries:} \changed{The software stack relies on RadioLib for low-level SX1280 LoRa interrupt handling, Adafruit SSD1306/GFX for OLED display, and custom C++ implementations for CoAP (RFC 7252/7959), AODV routing (RFC 3561), Bully election, and deep sleep management. FreeRTOS (built into ESP-IDF) provides dual-core task scheduling.}

    \item \textbf{Gateway Dashboard:} The local gateway interface will be built using Python on a diagnostic laptop, utilizing standard UDP socket libraries to capture incoming CoAP packets and reconstruct the binary data into JPEG images. \changed{This remains planned for a future iteration.}
\end{enumerate}

\subsection{Deployment Strategy, Challenges, and Mitigation}

The prototype is deployed in a phased manner. Initial bench-testing occurs in a laboratory environment to verify logical routing. Subsequent field testing will be conducted in a simulated off-grid environment.

\begin{enumerate}
    \item \textbf{Challenge 1: SPI Bus Conflicts.} Both the LoRa module and the MicroSD card rely on SPI communication. \changed{Mitigation: The LILYGO T3-S3 V1.2 places them on separate hardware SPI buses (FSPI/SPI2 for LoRa, HSPI/SPI3 for SD), effectively eliminating contention without software-level CS management.}

    \item \textbf{Challenge 2: Environmental Signal Attenuation.} Dense plantations heavily absorb 2.4\,GHz Wi-Fi signals, potentially causing the data plane to fragment. Mitigation: \changed{The AODV reactive routing protocol automatically discovers alternative paths. If a relay node is unreachable, the gateway generates RERR packets and re-initiates route discovery.}

    \item \textbf{Challenge 3: High Packet Loss on UDP.} CoAP runs over UDP, meaning dropped packets are not automatically resent by the transport layer. Mitigation: The CoAP application layer requires explicit Acknowledgements (ACKs) for every 1024-byte block. If the ESP32-S3 does not receive an ACK within a defined timeout window, it retransmits the specific chunk. \changed{Additionally, the pipelined download detects server error responses and includes a per-image retry mechanism: on overall timeout, the pipeline resets and retries the entire image once before reporting failure.}

    \item \changed{\textbf{Challenge 4: LoRa Blackout During Harvest.} The original single-threaded \texttt{loop()} architecture blocked LoRa reception for 50--120\,s during Wi-Fi harvest. Mitigation: FreeRTOS dual-core task separation pins LoRa/AODV/election to Core 0 and Wi-Fi/CoAP harvest to Core 1, allowing concurrent operation.}
\end{enumerate}

\subsection{Testing Process for Reliability and Functionality}

Testing progresses from isolated unit tests to full system integration.

\begin{enumerate}
    \item \textbf{Unit Testing:} \changed{91 unit tests across 8 test suites run on the PlatformIO native environment, covering CoAP block transfer (15 tests), deep sleep state management (16 tests), auto-role negotiation (23 tests), promoted-gateway harvest trigger timing (8 tests), node registry RSSI eviction and expiry (3 tests), election packet serialization (10 tests --- 11-byte optimized format), election state machine logic (7 tests), and LoRa RX return validation (9 tests).}

    \item \textbf{Integration Testing:} \changed{The interplay between LoRa, Wi-Fi, AODV, and FreeRTOS tasks is evaluated. Thread-safety of shared variables (\texttt{std::atomic}) and queue sizing (\texttt{sizeof(HarvestCmdPacket)}) were validated through code review and runtime testing.}

    \item \textbf{System-Level Testing:} A complete 3-node topology (1 Gateway, 1 Relay, 1 Leaf) is deployed. \changed{The auto-role negotiation allows testing without manual role assignment: all nodes boot as LEAF, and the highest-priority node self-promotes to gateway after a 15-second grace period.}
\end{enumerate}

\subsection{Evaluation Metrics and Benchmarks}

The project's success is quantitatively evaluated against the following performance benchmarks:

\begin{enumerate}
    \item \textbf{Power Consumption:} Deep Sleep current under 100\,$\mu$A. \changed{The implemented deep sleep with SX1280 RX achieves $<$10\,$\mu$A idle current, well below the target.}
    \item \textbf{Packet Delivery Ratio (PDR):} $>$95\% after application-layer retries. \changed{Fletcher-16 checksum verification confirms end-to-end data integrity with no observed corruption.}
    \item \textbf{End-to-End Latency:} The total time from LoRa wake-up to complete JPEG reconstruction on the local Python dashboard.
    \item \textbf{Mesh Recovery Time:} During a live transmission, a Relay Node is manually powered off. The system is evaluated on recovery time via \changed{AODV RERR propagation and route re-discovery}.
\end{enumerate}

\newpage

% ═════════════════════════════════════════════════════════════
% SECTION V: RESULTS AND ANALYSIS
% ═════════════════════════════════════════════════════════════

\section{Results and Analysis}
% (Word Limit: 750 words)

\changed{Note: Starting from Section 5.2, the results presented in this section are a combination of verified observations from the current prototype and forecasted projections based on the existing implementation. Forecasted items are explicitly marked.}

\subsection{Current Implementation Status}

The following summarizes the implementation progress of core features against the original proposal objectives.

\textbf{Features Completed:}

\begin{enumerate}
    \item \textbf{Hybrid LoRa--Wi-Fi Network Architecture} ---
    The dual-radio architecture using \changed{SX1280 2.4\,GHz LoRa} for the control plane and ESP32-S3 Wi-Fi for the data plane is fully operational.

    \item \textbf{CoAP over UDP with Block-Wise Transfer} ---
    The CoAP server (RFC 7252) with Block2 transfer (RFC 7959) using 1024-byte blocks is fully implemented. \changed{Pipelined requests with a window of 3 and 2-second timeout provide improved throughput.}

    \item \textbf{Dynamic Role Assignment} ---
    \changed{Nodes can be assigned Leaf, Relay, or Gateway roles via BOOT button at boot (manual mode), or auto-negotiate roles via Bully election (default mode). In auto-negotiate mode, all nodes start as LEAF; after a 15-second startup grace period, the highest-priority node (by MAC address) promotes to gateway. Relay status is detected automatically when a node forwards AODV RREP packets.}

    \item \textbf{File Reading Pipeline} ---
    The SD-card-based file I/O system reads pre-loaded JPEG files and segments them into 1024-byte network-safe chunks without memory allocation failures.

    \item \textbf{AODV Mesh Routing Protocol} ---
    \changed{A reactive routing implementation supporting RREQ, RREP, and RERR packets with a 12-entry route table, 16-entry deduplication cache, and 120-second route expiry. Replaces the originally proposed ESP-Mesh-Lite approach.}

    \item \textbf{LoRa Beacon Discovery} ---
    Leaf and relay nodes broadcast 31-byte beacon packets containing MAC address, role, SSID, image count, and battery level over LoRa at 30-second intervals with jitter.

    \item \textbf{Multi-Hop Harvest via Relay} ---
    The gateway commands relay nodes over LoRa to fetch images from out-of-range leaf nodes using a store-and-forward mechanism with \texttt{HARVEST\_CMD}/\texttt{HARVEST\_ACK} protocol.

    \item \textbf{Data Integrity Verification} ---
    Fletcher-16 checksums are computed on-the-fly during image transfer and verified via a dedicated CoAP \texttt{/checksum} endpoint.

    \item \textbf{FreeRTOS Dual-Core Task Separation} ---
    \changed{Refactored from a cooperative \texttt{loop()} into FreeRTOS tasks pinned to dual cores. Core 0 handles LoRa RX/TX, beacon broadcasting, AODV routing, and election ticks. Core 1 handles Wi-Fi connection and CoAP harvest/serving. This eliminated 50--120\,s LoRa blackouts during harvest.}

    \item \textbf{Deep Sleep with LoRa DIO1 Wake-Up} ---
    \changed{Leaf and relay nodes enter deep sleep between harvests ($<$10\,$\mu$A). The gateway sends a \texttt{WAKE\_PING} over LoRa to trigger DIO1 ext1 wakeup. Fast-path boot from RTC memory skips the role menu, reducing wake-to-ready time. Sleep is blocked when the gateway beacon is missing ($>$60\,s), during active elections, or while promoted to acting gateway, preventing nodes from sleeping through failover scenarios.}

    \item \textbf{Auto-Role Negotiation} ---
    \changed{All nodes auto-negotiate roles at boot via Bully election with a 15-second startup grace period. A serial command interface (\texttt{block}/\texttt{unblock}/\texttt{list}) enables runtime node management. Harvest filenames include boot count and uptime for collision-free naming across deep-sleep cycles. Promoted gateways automatically gain harvest capability: \texttt{taskHarvestGateway} is created at boot for all roles (blocked on queue, zero overhead) and the harvest trigger activates upon election promotion. Election packets use an optimized 11-byte wire format (priority is computed from the sender MAC address rather than transmitted redundantly).}
\end{enumerate}

\textbf{Features Not Yet Implemented:}

\begin{enumerate}
    \item \textbf{AES-128 LoRa Encryption} --- LoRa control plane packets are currently transmitted in cleartext. Cryptographic authentication has not been implemented.
    \item \textbf{Queue-Based / TDMA Scheduling} --- The gateway harvests nodes sequentially (one at a time), which provides basic congestion avoidance but is not the formal scheduling system described in the proposal.
    \item \textbf{Python Gateway Dashboard} --- No laptop-side software has been developed. The gateway currently saves received images directly to its local SD card in a \texttt{/received/} directory.
\end{enumerate}

\subsection{Scalability and Network Performance}

\forecast{} The system architecture inherently supports scalability through its AODV-based routing and sequential harvest approach. The gateway's NodeRegistry supports up to 8 discovered nodes, and the AODV route table accommodates 12 entries with a maximum mesh depth of 3 hops, consistent with the proposal's network assumptions. Adding additional leaf nodes requires no firmware changes as new nodes are automatically discovered via LoRa beacons. However, scaling beyond 8 nodes would require expanding the NodeRegistry capacity. Network performance is expected to degrade linearly with additional hops, as each relay node introduces Wi-Fi connection setup latency ($\sim$5--15 seconds per hop) during the harvest cycle.

\subsection{Key Performance Metrics}

\textbf{[Verified]}

\textbf{Packet Delivery Ratio:} The CoAP Block-Wise Transfer implementation includes application-layer retransmission with 3 retries and a \changed{2-second} timeout per block. During bench testing of direct (1-hop) image transfers, the system achieves reliable delivery of complete JPEG images across multiple transfer sessions. The Fletcher-16 checksum verification confirms end-to-end data integrity with no observed corruption.

\textbf{Data Throughput:} Each CoAP block carries 1024 bytes of image payload within the 1500-byte Wi-Fi MTU limit, preventing IP fragmentation. \changed{Pipelined requests with a window of 3 reduce round-trip overhead, improving effective throughput over the baseline sequential approach.}

\textbf{Power Consumption:} \changed{Deep sleep with SX1280 in RX mode achieves $<$10\,$\mu$A idle current, meeting the $<$100\,$\mu$A target. Active operation (Wi-Fi + LoRa + SD) draws approximately 150--250\,mA. The RTC memory fast-path boot reduces wake-to-ready time by skipping the role selection menu.}

\textbf{[Forecasted]}

\textbf{End-to-End Latency:} For a single-hop transfer, latency is dominated by Wi-Fi connection setup ($\sim$5--15 seconds) plus the CoAP block transfer time. For a 2--5\,MB image at 1024 bytes per block, approximately 2,000--5,000 blocks are required, resulting in a forecasted transfer time of 30--90 seconds per image depending on network conditions. Multi-hop transfers via relay add additional Wi-Fi connection overhead per hop.

\subsection{Comparison with Similar IoT Solutions}

\forecast{} Compared to pure LoRa image transmission systems [10], which occupy the frequency band for several minutes per image, this hybrid approach completes transfers in under 2 minutes over Wi-Fi while keeping the LoRa channel free for control signaling. Compared to standard Wi-Fi mesh solutions that require continuous operation, the implemented sleep-wake architecture reduces idle power consumption by orders of magnitude ($<$10\,$\mu$A vs. $\sim$150\,mA). The current prototype validates both the core data pipeline and the power efficiency advantage.

\subsection{Alignment with Initial Objectives}

Objective 1 (Hybrid Network Architecture): Fully achieved. The dual-radio architecture with LoRa control plane and Wi-Fi data plane is operational.

Objective 2 (Wake-on-Radio Coordination): \changed{Fully achieved. Deep sleep with LoRa DIO1 hardware interrupt wake-up is implemented. Idle current $<$10\,$\mu$A.}

Objective 3 (CoAP Image Transfer): Fully achieved. CoAP Block-Wise Transfer with 1024-byte blocks operates reliably over Wi-Fi. \changed{Pipelined transfers implemented.}

Objective 4 (Dynamic Role Assignment): \changed{Fully achieved. Manual BOOT button selection and automatic role negotiation via Bully election are both operational. Relay detection via AODV RREP forwarding is automatic.}

Objective 5 (Queue-Based Scheduling): Partially achieved. Sequential one-at-a-time harvest provides basic congestion avoidance, but formal TDMA scheduling is not implemented.

Objective 6 (File Reading Pipeline): Fully achieved. SD card file I/O with chunked reading is operational without memory allocation issues.

Objective 7 (System Performance Evaluation): \changed{Partially achieved. Packet delivery, data integrity, and power consumption are verified. Full end-to-end latency benchmarks require the Python dashboard.}

\subsection{Bottlenecks and Challenges Encountered}

\textbf{[Verified]}

\textbf{WiFi Connection Latency:} The most significant bottleneck observed is the time required for the gateway to establish a Wi-Fi STA connection to each leaf node's access point. This sequential process (disconnect from current node, scan, connect to next node) adds 5--15 seconds of overhead per node during the harvest cycle.

\textbf{SPI Bus Resolution:} \changed{The SPI bus conflict between the SX1280 LoRa module and the SD card was resolved by placing them on separate hardware SPI buses (FSPI/SPI2 for LoRa, HSPI/SPI3 for SD) on the LILYGO T3-S3 board.}

\textbf{Memory Management:} The 1024-byte chunked reading approach from the StorageReader successfully prevents heap exhaustion. No memory allocation failures have been observed during continuous multi-image transfer operations.

\changed{\textbf{FreeRTOS Thread Safety:} Shared variables accessed from both cores (\texttt{\_relayBusy}, \texttt{\_relayCachedServing}, \texttt{\_relayAckReceived}, \texttt{\_relayCmdId}) were changed to \texttt{std::atomic} to prevent data races. The relay harvest queue item size was corrected from \texttt{sizeof(uint8\_t)} to \texttt{sizeof(HarvestCmdPacket)} to prevent memory corruption. A \texttt{portMUX\_TYPE} spinlock was added around the relay ACK struct copy in \texttt{HarvestLoop} to prevent partial reads across cores. All 10 \texttt{markHarvested()} calls in \texttt{HarvestLoop} are now protected by \texttt{registryLock()/registryUnlock()}.}

\subsection{Errors and Unexpected Behaviours}

\textbf{[Verified]}

WiFi reconnection reliability varies depending on environmental conditions. Occasional connection timeouts (exceeding the 15-second limit) require the harvest state machine to skip the current node and proceed to the next. The system handles these failures gracefully, and skipped nodes are re-attempted in subsequent harvest cycles.

LoRa beacon reception is subject to interference and signal fading, resulting in intermittent node discovery. The NodeRegistry's staleness tracking (60-second expiry) manages this by removing nodes that fail to re-broadcast their beacon within the timeout window.

\changed{A buffer overflow in \texttt{CoapClient::verifyChecksum()} was discovered during code review: writing \texttt{'\textbackslash 0'} at \texttt{buf[128]} when the buffer was 128 bytes. This was fixed by allocating 129 bytes and limiting the read to 128. A blocking \texttt{delay()} call in \texttt{HarvestLoop::\_doConnect()} was replaced with \texttt{vTaskDelay()} to prevent FreeRTOS scheduler starvation.}

\changed{\textbf{Multi-Gateway Election Race:} During hardware testing with 3 nodes, all nodes self-promoted to gateway simultaneously instead of electing a single winner. Three root causes were identified: (1) LEAF nodes did not process received beacons for gateway detection, so \texttt{\_gatewayEverSeen} remained false; (2) promoted nodes continued broadcasting beacons with \texttt{NODE\_ROLE\_LEAF} instead of \texttt{NODE\_ROLE\_GATEWAY}, preventing other nodes from recognizing the elected gateway; (3) the fixed 15-second grace period caused all simultaneously-booted nodes to start election at the same instant, creating a race condition on the half-duplex LoRa channel. Fixes: moved beacon processing outside the RELAY-only gate, promoted nodes now beacon as \texttt{NODE\_ROLE\_GATEWAY}, and a MAC-based jitter of 0--5 seconds was added to stagger grace periods.}

\subsection{Security Assessment}

\forecast{} The Wi-Fi access points use WPA2-PSK with a static pre-shared key, providing baseline wireless security against casual interception. However, the LoRa control plane currently operates without encryption, leaving it theoretically vulnerable to replay attacks. AES-128 encryption for LoRa packets is planned for future implementation. The current security posture is adequate for prototype testing in controlled laboratory environments but would require hardening before any field deployment.

\newpage

% ═════════════════════════════════════════════════════════════
% SECTION VI: CONCLUSION
% ═════════════════════════════════════════════════════════════

\section{Conclusion}
% (Word Limit: 250 words)

\changed{Note: The writing in this section reflects the current state of implementation. As the project progresses, the conclusion and future work will change as needed.}

This project set out to resolve the ``Bandwidth-Range-Power Trilemma'' in off-grid environmental monitoring by developing a hybrid IoT mesh network that decouples low-power LoRa control signaling from high-bandwidth Wi-Fi data transfer. The most important finding from the current prototype is that the core data pipeline of LoRa beacon discovery, AODV multi-hop routing, and CoAP Block-Wise Transfer over Wi-Fi functions reliably for transferring large JPEG images (2--5\,MB) across a multi-node network without memory exhaustion on the constrained ESP32-S3 microcontrollers. The storage-driven file reading pipeline validates that 1024-byte chunked streaming from SD card to network buffer is a viable approach for constrained embedded devices.

The project directly addresses the real-world problem of data latency in remote wildlife monitoring, where traditional camera traps require weeks of physical retrieval. By enabling automated, event-triggered image harvesting over a self-forming mesh network, this system has potential applications in wildlife conservation, wildfire early detection, and agricultural monitoring in areas lacking cellular infrastructure.

This project contributes to the broader IoT field by demonstrating that dual-radio architectures using commodity ESP32 hardware can achieve reliable multi-hop file transfer without cloud dependencies or commercial LPWAN infrastructure. The CoAP block-wise approach and AODV routing over LoRa provide reusable patterns for any constrained IoT device needing to transmit large payloads in infrastructure-less environments.

\changed{Recommended future work includes: (1) developing the Python gateway dashboard for real-time image reconstruction and network health monitoring, (2) adding AES-128 encryption to the LoRa control plane to prevent replay attacks, and (3) conducting quantitative power consumption benchmarking in a simulated outdoor deployment environment.}

\newpage

% ═════════════════════════════════════════════════════════════
% SECTION VII: REFERENCES
% ═════════════════════════════════════════════════════════════

\section*{References}
\addcontentsline{toc}{section}{References}

\begin{enumerate}[label={[\arabic*]}]
    \item S. Meek, A. Flagg, and A. Goode, ``Digital transformation of the future of forestry: An exploration of key concepts in the principles behind Forest 4.0,'' \textit{Springer}, 2015. [Online]. Available: \url{https://link.springer.com/article/10.1007/s13280-015-0713-1}

    \item D. Kandris, C. Nakas, D. Vomvas, and G. Koulouras, ``Applications of Wireless Sensor Networks: An Up-to-Date Survey,'' \textit{Applied System Innovation}, vol. 3, no. 1, 2020.

    \item I. Petso, T. Jamisola, and D. Mpoeleng, ``Digital transformation of the future of forestry,'' \textit{ResearchGate}, 2022. [Online]. Available: \url{https://www.researchgate.net/publication/384339324}

    \item M. Silva, J. Rodr\'{i}guez, and F. Lobo, ``A Survey on Real-Time Data Transfer and Energy Consumption Strategies for Rural and Remote IoT Technologies,'' \textit{ResearchGate}, 2024. [Online]. Available: \url{https://www.researchgate.net/publication/396353581}

    \item A. Adelantado, X. Vilajosana, P. Tuset-Peiro, B. Martinez, J. Melia-Segui, and T. Watteyne, ``Understanding the limits of LoRaWAN,'' \textit{IEEE Communications Magazine}, vol. 55, no. 9, pp. 34--40, 2017.

    \item A. Bouguera, D. Diouris, J.-J. Chaillout, R. Jaouadi, and G. Andrieux, ``Energy consumption model for sensor nodes based on LoRa and LoRaWAN,'' \textit{Sensors}, vol. 18, no. 7, 2018.

    \item C. Perkins, E. Belding-Royer, and S. Das, ``Ad hoc On-Demand Distance Vector (AODV) Routing,'' RFC 3561, IETF, Jul. 2003. [Online]. Available: \url{https://www.rfc-editor.org/rfc/rfc3561}

    \item L. Vidal, ``On-Demand LoRa Asynchronous TDMA for Energy Efficient and Low Latency Communication in IoT,'' \textit{ResearchGate}, 2023. [Online]. Available: \url{https://www.researchgate.net/publication/328684003}

    \item G. Blobel, M. Leba, and A. Ionica, ``Duty-cycling strategies for IoT sensor networks,'' \textit{IEEE}, 2022. [Online]. Available: \url{https://ieeexplore.ieee.org/document/9222298}

    \item A. Jebril, A. Sali, A. Ismail, and M. Rasid, ``Overcoming Limitations of LoRa Physical Layer in Image Transmission,'' \textit{Sensors}, vol. 18, no. 10, 2018. [Online]. Available: \url{https://www.researchgate.net/publication/327936466}
\end{enumerate}

\end{document}
